---
title: "The Rise of the Three Point Shot"
font: 12pt
author: "Owen Schuler"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
Sys.setenv("VROOM_CONNECTION_SIZE" = 500000)
library(tidyverse)
library(ggplot2)
library(nbastatR)
library(png)
library(gganimate)
library(jpeg)
```

```{r, include=FALSE} 
yearkey <-c(seq(2005,2024,1)) #Creates year vector
nug_shots <- data.frame() #Initializes data frame

for (i in yearkey) {
  nug1 <- teams_shots(teams = "Denver Nuggets", seasons = i) #loops through all years
  if (nrow(nug1) > 0) { #Confirms data is present
    nug2 <- nug1 %>% 
      filter(isShotAttempted == TRUE) %>% #filters out unattempted shots
      select(yearSeason, slugSeason, typeEvent, typeAction, typeShot, zoneBasic, zoneRange, nameZone, slugZone, locationX, locationY, distanceShot) #selects necessary variables
    nug_shots <- bind_rows(nug_shots, nug2)  #stores in DF
  }
}
```

```{r,warning=FALSE}
fg_zones <- function(shots) { #creates function with shots parameter expecting shot data with locationX and Y variables
  court_img <- readPNG("Nug Court2.png")  # Reads in picture of court
  
  shots$is_make <- ifelse(shots$typeEvent == "Made Shot", 1, 0)  #creates a new column with a binary variable to indicate makes and misses
  make_percentage <- shots %>% #creates a new table with fg% by zone
    group_by(nameZone, yearSeason) %>% #groups by zone and by year (year for animation)
    summarise(make_pct = mean(is_make), .groups = "drop")  #takes average of binary column which is equal to the percentage of makes
  
  shots <- shots %>% #joins the fg% table by zone and year and removes NAs
    left_join(make_percentage, by = c("nameZone", "yearSeason")) %>% 
    na.omit() 
  
  y_out <- 855 #sets dimensions for court
  y_in <- -35
  x_out <- 255
  x_in <- -x_out #ensures symmetry
  
  ggplot(shots, aes(x = locationX, y = locationY)) + #uses shot data passed into the function with location data
    geom_point(aes(color = make_pct), alpha = 0.3, size = 3) + #makes the shots appear as points with size 3 and .3 opacity
    scale_color_gradient(low = "red", high = "green", limits = c(0,.6) ,labels= scales::percent_format(accuracy = 1)) + #creates a gradient color scale for shooting zones to indicate fg% as a continuous variable. capped at .6 so that differences between percentages are more pronounced.
    annotation_raster(court_img, xmin = x_in - 5, xmax = x_out + 5, ymin = y_in - 5, ymax = y_out + 15) + #stretches image to align with where 3pt vs 2pt shots as well as the basket are in the data.
    theme_void() +
    scale_x_continuous(limits = c(x_in, x_out)) + #sets limits for the court
    scale_y_continuous(limits = c(y_in, y_out)) +
    theme(aspect.ratio = (y_out - y_in) / (x_out - x_in)) + #ensures consistent ratio/scale
    labs(color = "Field Goal %", title = "Denver Nuggets Shots Hot Zones since 2005", subtitle ="{closest_state} Season")+ #labels + dynamic label for flipping through seasons
    transition_states(yearSeason, transition_length = 0, state_length = 1)+ 
    ease_aes('linear') #ensures constant speed for animation
}
#nug_shots_2023 <- nug_shots %>% filter(yearSeason == 2023) REMOVE
fg_gif <- animate(fg_zones(nug_shots), 
                fps = 1, 
                  nframes =  20, 
                width = 500, 
                height = 600, 
                  renderer = gifski_renderer()
                  ) #creates gif object by animating the function created beforehand
fg_gif
```




```{r,include=FALSE}
make_or_miss <- function(shots) {
  court_img <- readPNG("Nug Court2.png")  # Reads in picture of court
  
  shots$is_make <- ifelse(shots$typeEvent == "Made Shot", 1, 0)  #creates a new column with a binary variable to indicate makes and misses
 
  y_out <- 855
  y_in <- -35
  x_out <- 255
  x_in <- -x_out #sets dimensions for court
  
  ggplot(shots, aes(x = locationX, y = locationY)) + #uses shot data passed into the function with location data
    geom_point(aes(color = factor(is_make, levels = c(0,1), labels = c("Miss", "Make")), #correctly labels misses and makes
    shape = factor(is_make, levels = c(0,1), labels = c("Miss", "Make"))), #makes color and shape dependent on shot type
               alpha = 0.5, size = 2) + 
    scale_color_manual(values = c("Miss" = "#FEC524", "Make" = "#1D428A"))+
    scale_shape_manual(values = c("Miss" = 4, "Make" = 16))+  # Circles for makes, X's for misses
    annotation_raster(court_img, xmin = x_in - 5, xmax = x_out + 5, ymin = y_in - 5, ymax = y_out + 15) + #stretches image to align with where 3pt vs 2pt shots as well as the basket are in the data.
    theme_void() +
    scale_x_continuous(limits = c(x_in, x_out)) + #sets limits for the court
    scale_y_continuous(limits = c(y_in, y_out)) +
    theme(aspect.ratio = (y_out - y_in) / (x_out - x_in)) + #ensures consistent ratio/scale
    labs(color = NULL, title = "Denver Nuggets Shot Chart since 2005", subtitle = "{closest_state} Season") + #labels
    #transition_states(yearSeason, transition_length = 0, state_length = 1) +#separates transitions
    guides(
    color = guide_legend(title = NULL), 
    shape = guide_legend(title = NULL)) #+ #drops titles from legend
    #ease_aes('linear') #constant animation
}
#make <- make_or_miss(nug_shots)
#make
```



```{r,include=FALSE}
#make_gif <- animate(make_or_miss(nug_shots), 
                    #fps = 1, 
                    #nframes = 20, 
                    #width = 500, 
                    #height = 600, 
                    #renderer = gifski_renderer('make.gif')
                    #) #creates gif object by animating the function created beforehand
#make_gif
#make_or_miss(nug_shots)
#fg_zones(nug_shots)
```


# Analysis
- Since 2005, the Nuggets' shot selection has evolved to favor more 3 pointers and less mid range shots.
- As seen in the shot chart, as time passes, points become denser around the arc and inside the arc becomes sparse.
- The hot zones have also evolved, with above the break threes hitting at around 38% compared to around 33% in the earlier years.
- Further, of shots inside the arc, they have become concentrated within the key and also hitting at higher rates.

# In Depth
When looking at 3 pointers compared to mid range shots year over year, the ratio of 3 pointers to mid range shots has increased dramatically. In 2005, there was only one 3 pointer taken for every four mid range shots. This ratio jumped one 3 for every two mid range in 2007 and has steadily increased until 2018. From 2018 to 2024, the ratio has completely flipped with about two 3s being taken for every one mid range in each of the past 6 seasons. Further the proportion of shots taken from behind the 3 point line and in the paint have increased from 14% to 35% and 48% to 51% respectively while the proportion fo shots at mid range decreased from 38% to 14%. This shows how in this time, 3 point shooting has replaced the mid range shot. With the new prominence in 3 point shooting, above the break 3s have hit at 37-38% in recent years while only 33% in the 2000s. The data for left and right corner 3s is very noisy with much of the changes attributable to variance. Finally, shots in the paint have seen close to a 13 percentage point increase in field goal percentage since 2005 from ~36% to ~49%. Paint shots within the restricted area have gone in at around 67% on average over the past 5 years while only around 62% in the first 5 years. This could be attributable to the recent success of Nikola Jokic.

